### 2017-11-02 submission/manuscript.Rmd
add first paragraph of discussion

***Genera Previously Associated with Carcinoma Increases Relative Risk More than Alpha Diversity:***
Both fecal and tissue samples had a significantly increased RR for carcinoma but not for adenoma [Figure 3] which was greater than either evenness or Shannon diversity [Figure 2 & 3]. The relative risk did not increase when considering the total abundance or increasing number of carcinoma associated genera [Figure 3]. The RR effect size was greater for stool (RR range = `r paste(round(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`) then for tissue (RR range = `r paste(round(min(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`) . This decrease may be explained by the fact that tissue samples include matched samples.

***Using the Whole Community Increases Model AUC:***
For both fecal and tissue samples (matched and unmatched) there was a decrease in AUC when only OTUs from the CRC associated genera are used [Figure 4 & 5]. This decrease is observed in both adenoma and carcinoma groups [Figure 4 & 5]. The genus models generally had similar trends as observed for the OTU based models with the full genera models performing better then the CRC associated genera models [Figure S4-S5]. Both genus models perform similarily in their ability to be able to predict lesion (adenoma or carcinoma) with carcinoma having a higher AUC then adenoma [Figure S6-S8]. Matched tissue samples for those with carcinoma had an AUC that was more similar to the adenoma models [Figure S6A, S7B, & S8] then carcinoma models [Figure S6B & S7A].

@@

@@

>>>>

***Genera Previously Associated with Carcinoma Increases Relative Risk More than Alpha Diversity:***
Both fecal and tissue samples had a significantly increased RR for carcinoma but not for adenoma [Figure 3] which was greater than either evenness or Shannon diversity [Figure 2 & 3]. The relative risk did not increase when considering the total abundance or increasing number of carcinoma associated genera [Figure 3]. The RR effect size was greater for stool (RR range = `r paste(round(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`) then for tissue (RR range = `r paste(round(min(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`). This decrease may be explained by the fact that tissue samples include matched samples.

***Using the Whole Community Increases Model AUC over CRC Associated Genera:***
For both fecal and tissue samples (matched and unmatched) there was a decrease in AUC when only OTUs from the CRC associated genera are used [Figure 4 & 5]. This decrease is observed in both adenoma and carcinoma groups [Figure 4 & 5]. The genus models generally had similar trends as observed for the OTU based models with the full genera models performing better then the CRC associated genera models [Figure S4-S5]. Both genus models perform similarily in their ability to be able to predict lesion (adenoma or carcinoma) with carcinoma having a higher AUC then adenoma [Figure S6-S8]. Matched tissue samples for those with carcinoma had an AUC that was more similar to the adenoma models [Figure S6A, S7B, & S8] then carcinoma models [Figure S6B & S7A].

@@

Our study identifies clear diversity changes both at the community level and within individual genera that are present in indivdiuals with carcinoma versus those without the disease. Although there was a step wise decrease in diversity from control to adenoma to carcinoma; this did not translate into large effect sizes for the relative risk of either of these two conditions. These clear changes were not easily recapitulated in those with adenoma. Even though CRC associated genera increase the relative risk of carcinoma they do not increase the relative risk of adenoma. This information suggests that these specific genera may not be the primary members of the microbial community that contributes to the formation of an adenoma but is for a carcinoma. Additionally, our data shows that by using the whole community our models perform better then when they only use the CRC associated genera. CRC associated genera are clearly important to carcinoma but the context or community in which these microbes are a part of can drastically increase the ability of models to make predictions. This data supports the concept that small localized changes within the community may be occuring that are important in the disease progression of colorectal cancer and that they may not directly involve CRC associated genera.

- Flynn review and how data supports an idea of early specific localized community changes that becomes global in CRC

- Builiding on this idea moving forward how CRC associated genera interact with the overall community rather than in isolation will be important since they do not increase the RR of polyps
  - suggesting localized changes to community may be more important

- In closing to get at these more difficult questions the Power calculations and effect size in future studies need to be accounted for especially for adenoma where the predicted effect sizes are going to be small

@@

-aggregate studies done with feces and stool to investigate the role of the microbiota in adenoma and carcinoma.

- Significantly adds to the existing research using this approach by expanding both the total individuals included and the scope.

- leads to clear identifications of the challenges that lie ahead with respect to adenoma

### 2017-11-02 submission/manuscript.tex
add first paragraph of discussion

\textbf{\emph{Genera Previously Associated with Carcinoma Increases
Relative Risk More than Alpha Diversity:}} Both fecal and tissue samples
had a significantly increased RR for carcinoma but not for adenoma
{[}Figure 3{]} which was greater than either evenness or Shannon
diversity {[}Figure 2 \& 3{]}. The relative risk did not increase when
considering the total abundance or increasing number of carcinoma
associated genera {[}Figure 3{]}. The RR effect size was greater for
stool (RR range = 1.78 - 2.64) then for tissue (RR range = 1.33 - 1.53)
. This decrease may be explained by the fact that tissue samples include
matched samples.

\textbf{\emph{Using the Whole Community Increases Model AUC:}} For both
fecal and tissue samples (matched and unmatched) there was a decrease in
AUC when only OTUs from the CRC associated genera are used {[}Figure 4
\& 5{]}. This decrease is observed in both adenoma and carcinoma groups
{[}Figure 4 \& 5{]}. The genus models generally had similar trends as
observed for the OTU based models with the full genera models performing
better then the CRC associated genera models {[}Figure S4-S5{]}. Both
genus models perform similarily in their ability to be able to predict
lesion (adenoma or carcinoma) with carcinoma having a higher AUC then
adenoma {[}Figure S6-S8{]}. Matched tissue samples for those with
carcinoma had an AUC that was more similar to the adenoma models
{[}Figure S6A, S7B, \& S8{]} then carcinoma models {[}Figure S6B \&
S7A{]}.

@@

@@

>>>>

\textbf{\emph{Genera Previously Associated with Carcinoma Increases
Relative Risk More than Alpha Diversity:}} Both fecal and tissue samples
had a significantly increased RR for carcinoma but not for adenoma
{[}Figure 3{]} which was greater than either evenness or Shannon
diversity {[}Figure 2 \& 3{]}. The relative risk did not increase when
considering the total abundance or increasing number of carcinoma
associated genera {[}Figure 3{]}. The RR effect size was greater for
stool (RR range = 1.78 - 2.64) then for tissue (RR range = 1.33 - 1.53).
This decrease may be explained by the fact that tissue samples include
matched samples.

\textbf{\emph{Using the Whole Community Increases Model AUC over CRC
Associated Genera:}} For both fecal and tissue samples (matched and
unmatched) there was a decrease in AUC when only OTUs from the CRC
associated genera are used {[}Figure 4 \& 5{]}. This decrease is
observed in both adenoma and carcinoma groups {[}Figure 4 \& 5{]}. The
genus models generally had similar trends as observed for the OTU based
models with the full genera models performing better then the CRC
associated genera models {[}Figure S4-S5{]}. Both genus models perform
similarily in their ability to be able to predict lesion (adenoma or
carcinoma) with carcinoma having a higher AUC then adenoma {[}Figure
S6-S8{]}. Matched tissue samples for those with carcinoma had an AUC
that was more similar to the adenoma models {[}Figure S6A, S7B, \& S8{]}
then carcinoma models {[}Figure S6B \& S7A{]}.

@@

Our study identifies clear diversity changes both at the community level
and within individual genera that are present in indivdiuals with
carcinoma versus those without the disease. Although there was a step
wise decrease in diversity from control to adenoma to carcinoma; this
did not translate into large effect sizes for the relative risk of
either of these two conditions. These clear changes were not easily
recapitulated in those with adenoma. Even though CRC associated genera
increase the relative risk of carcinoma they do not increase the
relative risk of adenoma. This information suggests that these specific
genera may not be the primary members of the microbial community that
contributes to the formation of an adenoma but is for a carcinoma.
Additionally, our data shows that by using the whole community our
models perform better then when they only use the CRC associated genera.
CRC associated genera are clearly important to carcinoma but the context
or community in which these microbes are a part of can drastically
increase the ability of models to make predictions. This data supports
the concept that small localized changes within the community may be
occuring that are important in the disease progression of colorectal
cancer and that they may not directly involve CRC associated genera.

\begin{itemize}
\item
  Flynn review and how data supports an idea of early specific localized
  community changes that becomes global in CRC
\item
  Builiding on this idea moving forward how CRC associated genera
  interact with the overall community rather than in isolation will be
  important since they do not increase the RR of polyps
\item
  suggesting localized changes to community may be more important
\item
  In closing to get at these more difficult questions the Power
  calculations and effect size in future studies need to be accounted
  for especially for adenoma where the predicted effect sizes are going
  to be small
\end{itemize}

@@

-aggregate studies done with feces and stool to investigate the role of
the microbiota in adenoma and carcinoma.

\begin{itemize}
\item
  Significantly adds to the existing research using this approach by
  expanding both the total individuals included and the scope.
\item
  leads to clear identifications of the challenges that lie ahead with
  respect to adenoma
\end{itemize}
